---
title: "lai_2026_figures_markdown"
output: html_document
date: "2026-04-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#final scripts for figures in LAI neolithic paper 2026 submission
#main paper figures:

#data found in final_figures_data direcotry 


#Figure 1:

```{r}

# Figure 1 

library(ggplot2)
library(dplyr)
library(readxl)
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)
library(patchwork)
library(scales)


# -------- Load ancestry + date data --------

dates <- read.table(
  "../final_figures_data/allentoft_dates_041725.txt",
  header = TRUE,
  sep = "\t",
  stringsAsFactors = FALSE
)

ancestry <- read.table(
  "../final_figures_data/qpadm_output_global_ancestry.txt",
  header = TRUE
)

colnames(dates) <- c("ID", "Age")
colnames(ancestry) <- c("ID", "Ancestry")

dates$Age <- as.numeric(gsub(",", "", dates$Age))

main_data <- merge(dates, ancestry, by = "ID")
main_data$Group <- "ADMIXED"


# -------- Reference panel --------

panel <- read.table(
  "../final_figures_data/flare.ref.panel",
  header = FALSE,
  stringsAsFactors = FALSE
)
colnames(panel) <- c("ID", "Label")

panel_with_dates <- merge(panel, dates, by = "ID")

panel_with_dates$Ancestry <- ifelse(panel_with_dates$Label == "neo", 0,
                                    ifelse(panel_with_dates$Label == "hg", 1, NA))

panel_with_dates$Group <- ifelse(panel_with_dates$Label == "neo", "FARMER",
                                 ifelse(panel_with_dates$Label == "hg", "HG", NA))

panel_with_dates <- panel_with_dates[!is.na(panel_with_dates$Group), ]


# -------- Combine --------

combined <- rbind(
  main_data[, c("ID", "Age", "Ancestry", "Group")],
  panel_with_dates[, c("ID", "Age", "Ancestry", "Group")]
)

admixed_data <- combined %>% filter(Group == "ADMIXED")


# -------- Map sample metadata --------

allentoft_excel_data <- read_excel(
  "../final_figures_data/allentoft_full_data.xlsx",
  skip = 10
)

admixed_samples <- read.table(
  "../final_figures_data/mneo_keep_fam.txt"
)$V2

source_samples <- read.table(
  "../final_figures_data/flare.ref.panel"
)

neo_samples <- source_samples$V1[source_samples$V2 == "neo"]
hg_samples  <- source_samples$V1[source_samples$V2 == "hg"]

filtered_data <- allentoft_excel_data %>%
  filter(`Individual ID` %in% c(admixed_samples, neo_samples, hg_samples)) %>%
  mutate(
    Group = case_when(
      `Individual ID` %in% admixed_samples ~ "ADMIXED",
      `Individual ID` %in% neo_samples ~ "FARMER",
      `Individual ID` %in% hg_samples ~ "HG"
    ),
    latitude = as.numeric(Latitude),
    longitude = as.numeric(Longitude)
  ) %>%
  filter(!is.na(latitude))


# -------- World map --------

world <- ne_countries(scale = "medium", returnclass = "sf")



# Define purple gradient
purple_gradient <- scales::gradient_n_pal(c("#dcd6f7", "#4b0082"))

# Scale and invert age for color gradient
admixed_data <- admixed_data %>%
  mutate(
    Age_scaled = (Age - min(Age)) / (max(Age) - min(Age)),
    Age_scaled_inv = 1 - Age_scaled,
    color_by_age = purple_gradient(Age_scaled_inv)
  )

# Stronger colors for groups
colors <- c(
  "FARMER" = "#E69F00",
  "HG"     = "#009E73"
)

# Merge age into map data
filtered_data <- filtered_data %>%
  left_join(admixed_data %>% select(ID, Age),
            by = c("Individual ID" = "ID")) %>%
  mutate(
    shape_final = case_when(
      Group == "FARMER" ~ 15,
      Group == "HG" ~ 17,
      Group == "ADMIXED" ~ 16,
      TRUE ~ 16
    )
  )

# Base theme
base_theme <- theme_classic() + 
  theme(
    axis.title = element_text(size = 11),
    axis.text = element_text(size = 10),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 9, face = "plain")
  )


# -------- Panel A --------

p1 <- ggplot(data = world) +
  geom_sf(fill = "gray80", color = "white") +
  
  # FARMER
  geom_point(
    data = filtered_data %>% filter(Group == "FARMER"),
    aes(x = longitude, y = latitude, shape = "Farmer"),
    color = "#E69F00",
    size = 2,   alpha = 0.7 
  ) +
  
  # HG
  geom_point(
    data = filtered_data %>% filter(Group == "HG"),
    aes(x = longitude, y = latitude, shape = "Hunter-Gatherer"),
    color = "#009E73",
    size = 2,   alpha = 0.7 
  ) +
  
    # ADMIXED (gradient)
  geom_point(
    data = filtered_data %>% filter(Group == "ADMIXED"),
    aes(x = longitude, y = latitude, color = Age),
    shape = 16,
    size = 2,   alpha = 0.85 
  ) +
  
  
  scale_color_gradient(
  low = "#dcd6f7",
  high = "#4b0082",
  name = "Date (years BP)",
  guide = guide_colorbar(
    title.theme = element_text(size = 9, margin = margin(t = 25))
  )
) +
  
  # the legend for Farmer/HG
  scale_shape_manual(
  values = c("Farmer" = 15, "Hunter-Gatherer" = 17),
  name = NULL,
  guide = guide_legend(
    override.aes = list(
      color = c("#E69F00", "#009E73")
    )
  )
) +
  
  coord_sf(xlim = c(-11, 37), ylim = c(35, 61)) +
  xlab("Longitude") + ylab("Latitude") +
  base_theme +
theme(
  legend.position = "right",
  legend.box = "vertical"
)
  


# -------- Panel B --------

p2 <- ggplot(admixed_data, aes(x = Age, y = Ancestry, color = Age)) +
  geom_point(size = 2,   alpha = 0.85) +
  scale_x_reverse() +
  scale_color_gradient(
    low = "#dcd6f7",
    high = "#4b0082",
    name = "Date (years BP)"
  ) +
  base_theme +
  labs(
    x = "Average Date (years BP)",
    y = "Hunter-Gatherer Ancestry"
  ) +
theme(
    legend.position = "none",
    plot.margin = margin(5, 5, 1, 5)  
  )


# -------- Combine + save --------

final_plot <- (p1 | p2) +
  plot_annotation(
    tag_levels = "A",
    theme = theme(
      plot.tag = element_text(face = "bold", size = 12),
      plot.margin = margin(0, 0, -20, 0)
    )
  ) &
  theme(plot.tag.position = c(0.06, 1.2))


ggsave(
  "Figure1.pdf",
  final_plot,
  width = 6.5,
  height = 3
)

```



#Figure 2:

```{r}

# Load required libraries
library(tidyverse)
library(ggplot2)
library(cowplot)



colors <- c(
  FARMER = "#808080",
  HG = "#808080",
  ADMIXED = "#9467bd",
  simplai = "#e377c2",
  ancestry_hmm = "#1f77b4",
  ancestry_paths = "#ff7f0e",
  mosaic = "#2ca02c",
  rfmix = "#d62728"
)


# Number of datasets
num_datasets <- as.numeric(7)

# Paths to each method's ancestry file
dataset_paths <- c("../final_figures_data/7v3_admixture.txt", "../final_figures_data/7v3_mosaic_global.txt", "../final_figures_data/ancestry_paths_output_global_ancestry.txt", "../final_figures_data/7v7_ancestry_hmm_global.txt", "../final_figures_data/7v7_rfmix_global_ancestry.txt","../final_figures_data/7v3_simplai_output_global_ancestry.txt", "../final_figures_data/qpadm_output_global_ancestry.txt")


# Method names
dataset_names <- c("imputed_admixture", "mosaic", "ancestry_paths", "ancestry_hmm", "rfmix", "simplai", "qpadm")

# Function to read and return data with method name
read_and_process <- function(file_path, method_name) {
  df <- read.table(file_path, header = TRUE)
  colnames(df) <- c("sample", paste0("global_ancestry_", method_name))
  return(df)
}

# Read all datasets into a list
dataset_list <- list()

for (i in 1:num_datasets) {
  dataset_list[[i]] <- read_and_process(dataset_paths[i], dataset_names[i])
}

# Combine all into one dataframe, spread by method
combined_df <- reduce(dataset_list, function(x, y) full_join(x, y, by = "sample"))

# Just select the columns we want (they already have the right names)
combined_df <- combined_df %>%
  select(sample, starts_with("global_ancestry_"))

# Drop any rows with NA values
combined_df <- na.omit(combined_df)

# A named vector to rename methods for plot titles
rename_methods <- c(
  imputed_admixture = "ADMIXTURE (7v3)",
  rfmix = "RFMix (7v7, 0.9)",
  simplai = "simpLAI (7v3)",
  mosaic = "Mosaic (7v3, 0.9)",
  ancestry_paths = "Ancestry Paths (7v48)",
  ancestry_hmm = "Ancestry HMM (7v7, 0.9)",
  qpadm = "qpAdm"
)

# Set base font size scaled to new size (was 16 before, let's reduce slightly)
base_font_size <- 10


plots <- list()
qpAdm_vals <- combined_df[["global_ancestry_qpadm"]]

for (method in dataset_names) {
  if (method != "qpadm") {
    y_vals <- combined_df[[paste0("global_ancestry_", method)]]

    df_plot <- data.frame(qpadm = qpAdm_vals, method_val = y_vals)

    r_val <- cor(df_plot$qpadm, df_plot$method_val, use = "complete.obs")
    r_text <- sprintf("R = %.2f", r_val)
    
    #delta:
     delta_val <- mean(abs(df_plot$method_val - df_plot$qpadm), na.rm = TRUE)
    delta_text <- sprintf("Δ = %.2f", delta_val)
    
        #slope of best-fit line
    fit <- lm(method_val ~ qpadm, data = df_plot)
    beta_val <- coef(fit)[2]
    beta_text <- sprintf("β = %.2f", beta_val)

    method_label <- rename_methods[[method]]
    if (is.null(method_label)) method_label <- tools::toTitleCase(gsub("_", " ", method))

    point_color <- if (method == "imputed_admixture") {
      colors["ADMIXED"]
    } else {
      colors[[method]] %||% "steelblue"
    }

    p <- ggplot(df_plot, aes(x = qpadm, y = method_val)) +
      geom_point(alpha = 0.8, color = point_color, size = 2) +  # adjust point size if needed
      geom_smooth(method = "lm", color = "firebrick", se = FALSE, linewidth = 1.2) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black", linewidth = 1) +
          annotate("text",
             x = 0.00, y = 0.60,
             label = bquote(R == .(round(r_val, 2))),
             hjust = 0, size = 4, fontface = "bold") +
    annotate("text",
             x = 0.00, y = 0.50,
             label = bquote(beta == .(round(beta_val, 2))),
             hjust = 0, size = 4, fontface = "bold") +
    annotate("text",
             x = 0.30, y = 0.08,
             label = bquote(Delta == .(round(delta_val, 2))),
             hjust = 0, size = 4, fontface = "bold") +
      
      labs(
        x = "qpAdm",
        y = "Global Ancestry",
        title = method_label
      ) +
      theme_minimal(base_size = base_font_size) +
      theme(
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey90", linewidth = 0.5),
        plot.title = element_text(face = "bold", size = base_font_size + 1, hjust = 0.5, margin = margin(b = 10)),
        axis.title = element_text(size = base_font_size + 2),
        axis.text = element_text(size = base_font_size, color = "black"),
        plot.background = element_rect(fill = "white", color = NA),
        panel.background = element_rect(fill = "white", color = NA)
      ) +
      coord_cartesian(xlim = c(0, 0.45), ylim = c(0, 0.7))

    plots[[length(plots) + 1]] <- p
  }
}

# Save plot grid as PDF with new size
final_plot <- plot_grid(plotlist = plots, ncol = 3, labels = NULL)

ggsave("Figure2.pdf", final_plot, width = 7, height = 5, dpi = 300)

```



#Figure 3
#note: I manually added text for each of the method labels, colors for each method follor from Figure 2

```{r}

#data analysis ran using: LA correlations with different window sizes:

#Rscript bins_correlation_matrix_names.R 6 ../rfmix/average_ancestry_output.txt ../ancestry_paths/average_ancestry_output.txt ../ancestry_hmm_haplotypes/average_ancestry_output.txt ../simplai/average_ancestry_output.txt ../mosaic/average_ancestry_output.txt /project/mathilab/gmies/neolithic_selection/allentoft_data/031825_rerun_filter_allen_LAI/selection_analysis/same_format_042125/replication/060225/7_9ploidy2_run_ancestry_hmm/output_ancestry_hmm.txt rfmix ancestry_paths ancestry_hmm simplai mosaic replication 10

#Rscript bins_correlation_matrix_names.R 6 ../rfmix/average_ancestry_output.txt ../ancestry_paths/average_ancestry_output.txt ../ancestry_hmm_haplotypes/average_ancestry_output.txt ../simplai/average_ancestry_output.txt ../mosaic/average_ancestry_output.txt /project/mathilab/gmies/neolithic_selection/allentoft_data/031825_rerun_filter_allen_LAI/selection_analysis/same_format_042125/replication/060225/7_9ploidy2_run_ancestry_hmm/output_ancestry_hmm.txt rfmix ancestry_paths ancestry_hmm simplai mosaic replication 1000



cor_10kb <- read.table("../final_figures_data/LA_correlation_matrix_binsize_10kb.txt")
cor_100kb <- read.table("../final_figures_data/LA_correlation_matrix_binsize_100kb.txt")
cor_1000kb <- read.table("../final_figures_data/LA_correlation_matrix_binsize_1000kb.txt")
cor_10000kb <- read.table("../final_figures_data/LA_correlation_matrix_binsize_10000kb.txt")

library(tidyverse)
library(cowplot)

#if less than 0, replace with 0:

cor_10000kb[cor_10000kb < 0] <- 0

# Store them in a list for easier looping
bin_sizes <- c(10, 100, 1000, 10000)
cor_list <- list("10" = cor_10kb, "100" = cor_100kb, "1000" = cor_1000kb, "10000" = cor_10000kb)


# Define focal methods
focal_methods <- c("AVG_rfmix", "AVG_mosaic", "AVG_simplai", "AVG_ancestry_hmm")


# Filter colors vector to just those methods that appear in correlation matrices
cor_methods <- c("rfmix", "ancestry_paths", "ancestry_hmm", "simplai", "mosaic", "replication")
colors_filtered <- colors[names(colors) %in% cor_methods]

# Helper function to extract correlation data for one focal method
build_corr_df <- function(focal_method) {
  # Remove the "AVG_" prefix to get the bare method name
  focal_name <- sub("^AVG_", "", focal_method)
  
  # Other methods are colors_filtered names except the focal method
  other_methods <- setdiff(names(colors_filtered), focal_name)
  
  result <- data.frame()
  
  for (size in bin_sizes) {
    cor_mat <- cor_list[[as.character(size)]]
    
    # Compose column names with AVG_ prefix
    req_cols <- paste0("AVG_", other_methods)
    
    # Check columns exist
    missing_cols <- setdiff(req_cols, colnames(cor_mat))
    if (length(missing_cols) > 0) {
      stop(paste("Missing columns:", paste(missing_cols, collapse = ", ")))
    }
    
    # Extract correlations for focal row vs others columns
    row_vals <- as.numeric(cor_mat[focal_method, req_cols])
    
    # Build data frame for this bin size
    df <- data.frame(
      Method = other_methods,
      Correlation = row_vals,
      BinSize = size,
      Focal = focal_name
    )
    
    result <- rbind(result, df)
  }
  
  return(result)
}

# Now combine all focal method data frames
all_data <- bind_rows(lapply(focal_methods, build_corr_df))


# Combine all focal method plots into one data frame
all_data <- bind_rows(lapply(focal_methods, build_corr_df))



# Capitalize and label methods for facet titles
label_map <- c(
  "rfmix" = "RFMix",
  "mosaic" = "Mosaic",
  "simplai" = "simpLAI",
  "ancestry_hmm" = "Ancestry HMM"
)


all_data$Focal <- factor(all_data$Focal, levels = names(label_map), labels = label_map)
all_data$Method <- factor(all_data$Method, levels = names(colors))



plot <- ggplot(all_data, aes(x = BinSize, y = Correlation, color = Method)) +
  geom_point(size = 2) +
  geom_line(linewidth = 1) +
  facet_wrap(~Focal, ncol = 2) +
  scale_color_manual(values = colors) +
  scale_x_log10(breaks = bin_sizes) +
  ylim(0, 0.8) +
  labs(x = "Bin Size (kb)", y = "Pearson Correlation") +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(size = 13, face = "bold"),
    legend.position = "none",
    panel.border = element_rect(color = "black", fill = NA),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 11),
    plot.margin = margin(5, 40, 5, 5),
    panel.spacing.x = unit(1, "lines")  # increase horizontal space between facets
  )

ggsave("Figure3.pdf", plot, width = 5, height = 4)

```



#Figure 4:
#note Figure 4 is made up of two panels that are created using different scripts:

#Code for Figure 4 A and B
#Figure 4 C D and E are created with tracts script 

#note: I edited by hand the positioning of LAI method labels.
```{r}


hg_tracts <- read.table("../final_figures_data/hg_combined_tract_data.txt", header = TRUE)
farmer_tracts <- read.table("../final_figures_data/farmer_combined_tract_data.txt", header = TRUE)


# Combined analysis for both populations
library(ggplot2)
library(dplyr)
library(gridExtra)
library(grid)
library(patchwork)  # Add this library for better plot combining

# Function to create plot data for a given population
create_plot_data <- function(pop_data, pop_name) {
  # Log-spaced bins from 1 to 1000 cM
  bin_edges <- 10^seq(log10(0.5), log10(1000), length.out = 60)
  
  # Filter out zeros and negative (tract length must be >0 for log binning)
  filtered_data <- pop_data %>%
    filter(tract_length_cm > 0)
  
  # Assign each tract to a bin
  filtered_data <- filtered_data %>%
    mutate(bin = cut(tract_length_cm, breaks = bin_edges, include.lowest = TRUE, right = FALSE))
  
  # Create a lookup table for bin centers
  bin_centers <- data.frame(
    bin = levels(cut(1, breaks = bin_edges, include.lowest = TRUE, right = FALSE)),
    bin_start = bin_edges[-length(bin_edges)],
    bin_end = bin_edges[-1]
  ) %>%
    mutate(bin_center = (bin_start + bin_end) / 2)
  
  # Population proportions
  p_farmer <- 0.6
  p_hg <- 0.4
  
  # Summarize counts per Source and bin
  binned_summary <- filtered_data %>%
    group_by(Source, bin) %>%
    summarise(count = n(), .groups = 'drop') %>%
    left_join(bin_centers, by = "bin") %>%
    filter(!is.na(bin_center))
  
  # Calculate total tracts per Source for empirical datasets
  total_counts_per_source <- binned_summary %>%
    group_by(Source) %>%
    summarise(total_count = sum(count))
  
  # Add proportion column for empirical data
  binned_summary <- binned_summary %>%
    left_join(total_counts_per_source, by = "Source") %>%
    mutate(proportion = count / total_count)
  
  # Define scaling factor based on population proportion
  p_pop <- ifelse(pop_name == "hg", p_hg, 
                  ifelse(pop_name == "farmer", p_farmer, NA))
  
  if (is.na(p_pop)) {
    stop("Population must be 'hg' or 'farmer' for theoretical scaling.")
  }
  
  # Generate continuous theoretical curves
  T_values <- seq(10, 200, by = 5)  # More continuous T values
  
  theoretical_data <- data.frame()
  
  # Define cutoff probability and calculate corresponding x-value for each T
  cutoff_prob <- 1e-5
  
  for (T in T_values) {
    lambda_rate <- (T * (1 - p_pop)) / 100
    
    # Calculate x-value where probability drops to cutoff_prob
    # For exponential: prob = exp(-lambda * x), so x = -log(prob) / lambda
    cutoff_x <- -log(cutoff_prob) / lambda_rate
    
    # Filter bin_centers to only include those up to cutoff_x
    filtered_bins <- bin_centers %>%
      filter(bin_center <= cutoff_x)
    
    # Add the exact cutoff point
    if(nrow(filtered_bins) > 0) {
      theory_bin_data <- filtered_bins %>%
        mutate(
          bin_probability = (exp(-lambda_rate * bin_start) - exp(-lambda_rate * bin_end)),
          Source = paste0("Theory T=", T),
          T_value = T
        )
      
      # Add the exact cutoff point
      cutoff_point <- data.frame(
        bin = NA,
        bin_start = cutoff_x,
        bin_end = cutoff_x,
        bin_center = cutoff_x,
        bin_probability = cutoff_prob,
        Source = paste0("Theory T=", T),
        T_value = T
      )
      
      theory_bin_data <- rbind(theory_bin_data, cutoff_point)
      theoretical_data <- rbind(theoretical_data, theory_bin_data)
    }
  }
  
  # Rename methods for colors
  change_color_names <- c(
    "rfmix" = "rfmix",
    "ancestry_paths" = "ancestry_paths",
    "ancestry_hmm_haplotypes" = "ancestry_hmm",
    "simplai" = "simplai",
    "mosaic" = "mosaic"
  )
  
  binned_summary$Source <- sapply(binned_summary$Source, function(x) {
    if (x %in% names(change_color_names)) {
      change_color_names[[x]]
    } else {
      x
    }
  })
  
  # Combine empirical and theoretical data
  empirical_data <- binned_summary %>% 
    select(Source, bin_center, proportion) %>% 
    rename(value = proportion) %>%
    mutate(type = "empirical")
  
  theoretical_plot_data <- theoretical_data %>% 
    select(Source, bin_center, bin_probability, T_value) %>% 
    rename(value = bin_probability) %>%
    mutate(type = "theoretical")
  
  # Filter theoretical data to only keep points with value >= 1e-5 (cutoff just below 0.00001)
  theoretical_plot_data <- theoretical_plot_data %>%
    filter(value >= 1e-5)
  
  list(empirical = empirical_data, theoretical = theoretical_plot_data)
}

# Create plot function
create_individual_plot <- function(plot_data_list, pop_name, y_limits, x_limits, show_x_title = TRUE, show_title = TRUE, point_shape = 16) {
  empirical_data <- plot_data_list$empirical
  theoretical_data <- plot_data_list$theoretical
  
  # Method names for labeling
  method_names <- c(
    "rfmix" = "RFMix",
    "ancestry_paths" = "Ancestry Paths", 
    "ancestry_hmm" = "Ancestry HMM",
    "simplai" = "SimpLAI",
    "mosaic" = "Mosaic"
  )
  
  # Get empirical sources
  empirical_sources <- unique(empirical_data$Source)
  
  # Create base plot
  x_title <- if(show_x_title) "Tract Length (cM)" else ""
  
  p <- ggplot() +
    # Add theoretical curves for selected T values
    geom_line(data = theoretical_data %>% filter(T_value %in% c(10, 20, 50, 100, 200)),
              aes(x = bin_center, y = value, group = T_value),
              color = "black", linewidth = 0.9, alpha = 0.8) +
    
    # Add empirical data points
    geom_point(data = empirical_data, 
               aes(x = bin_center, y = value, color = Source),
               shape = point_shape,
               alpha = 0.8, size = 1) +
    
    scale_x_log10(name = x_title, 
                  limits = x_limits,  # Use consistent x limits
                  breaks = c(1, 5, 20, 100, 500),
                  labels = c("1", "5", "20", "100", "500")) +
    scale_y_log10(name = "Proportion", 
                  limits = y_limits,
                  labels = scales::label_number(accuracy = 0.00001, trim = TRUE)) +
    
    scale_color_manual(values = colors[empirical_sources]) +
    theme_minimal() +
    theme(
      legend.position = "none",
      axis.title = element_text(size = 8),
      axis.text = element_text(size = 7),
      plot.margin = margin(t = 1, r = 15, b = -1, l = 3),
      panel.grid.minor = element_blank()
    )
  
  # Add T value labels at the end of each curve (find the maximum x for each T line)
  theory_label_data <- theoretical_data %>%
    filter(T_value %in% c(10, 20, 50, 100, 200)) %>%
    group_by(T_value) %>%
    filter(bin_center == max(bin_center)) %>%
    ungroup()
  
  p <- p + geom_text(data = theory_label_data,
                     aes(x = bin_center, y = value, label = paste0("T=", T_value)),
                     hjust = -0.05, vjust = 0.5, size = 2.3, color = "black")
  
  # Add method labels at the end of empirical data lines
  method_label_data <- empirical_data %>%
    group_by(Source) %>%
    filter(bin_center == max(bin_center)) %>%
    ungroup() %>%
    mutate(method_label = method_names[Source])
  
  p <- p + geom_text(data = method_label_data,
                     aes(x = bin_center, y = value, label = method_label, color = Source),
                     hjust = -0.05, vjust = 0.5, size = 2.3, show.legend = FALSE)
  
  # Add panel label only if requested
  if(show_title) {
    panel_label <- ifelse(pop_name == "hg", "Hunter-Gatherer", "Farmer")
    p <- p + ggtitle(panel_label) +
      theme(plot.title = element_text(hjust = 0, size = 7, face = "bold"))
  }
  
  return(p)
}

# Process both populations
hg_plot_data <- create_plot_data(hg_tracts, "hg")
farmer_plot_data <- create_plot_data(farmer_tracts, "farmer")

# Determine common y-axis and x-axis limits
all_empirical_values <- c(hg_plot_data$empirical$value, farmer_plot_data$empirical$value)
all_theoretical_values <- c(hg_plot_data$theoretical$value, farmer_plot_data$theoretical$value)
all_values <- c(all_empirical_values, all_theoretical_values)

# Set consistent limits
y_limits <- c(min(all_values) * 0.9, max(all_values) * 1.2)

# Get x-axis limits from both datasets
all_x_values <- c(hg_plot_data$empirical$bin_center, farmer_plot_data$empirical$bin_center,
                  hg_plot_data$theoretical$bin_center, farmer_plot_data$theoretical$bin_center)
x_limits <- c(min(all_x_values) * 0.9, max(all_x_values) * 1.1)

# Create individual plots with consistent limits
p_hg <- create_individual_plot(hg_plot_data, "hg", y_limits, x_limits, 
                              show_x_title = FALSE, show_title = FALSE, point_shape = 17)  # Triangle
p_farmer <- create_individual_plot(farmer_plot_data, "farmer", y_limits, x_limits, 
                                  show_x_title = TRUE, show_title = FALSE, point_shape = 15)  # Square

# Combine plots using patchwork
combined_plot <- p_hg / p_farmer
combined_plot <- combined_plot + 
  plot_layout(heights = c(1, 1)) +
  plot_annotation() &
  theme(plot.margin = margin(1, 2, 0.3, 2))

# Save the plot
ggsave("Figure4_AB.pdf", combined_plot, width = 3.5, height = 4)


```


#for Figure 4 C-E see: Figure4C-E_tractsplot.py
#data found in: final_figures_data/rfmix_tracts, mosaic_tracts, simplai_tracts


#Figure 5:

```{r}

# FULL FIGURE: Manhattan + QQ + Replication + Enrichment panels

library(ggplot2)
library(dplyr)
library(cowplot)
library(gridExtra)


# LOAD DATA

z_score_data <- read.table("../final_figures_data/dis_combined_weighted_z_scores.txt", header = TRUE)

capture_z_score_data <- read.table("../final_figures_data/rep1_combined_weighted_z_scores.txt", header = TRUE)

rep_z_score_data <- read.table("../final_figures_data/rep2_combined_weighted_z_scores.txt", header = TRUE)

chr23_data <- read.table(
  "../final_figures_data/chrx_combined_weighted_z_scores.txt",
  header = TRUE
)

combined_data <- bind_rows(z_score_data, chr23_data)
combined_data$CHR <- as.numeric(combined_data$CHR)


# EQUAL-WIDTH CHROMOSOMES

combined_data <- combined_data %>%
  group_by(CHR) %>%
  arrange(POS) %>%
  mutate(pos_scaled = (POS - min(POS)) / (max(POS) - min(POS))) %>%
  ungroup()

chr_offsets <- combined_data %>%
  distinct(CHR) %>%
  arrange(CHR) %>%
  mutate(offset = row_number() - 1)

combined_data <- combined_data %>%
  left_join(chr_offsets, by = "CHR") %>%
  mutate(plot_index = pos_scaled + offset)


# COLORS

combined_data$color <- ifelse(
  combined_data$CHR == 23,
  "#5dade2",
  ifelse(combined_data$CHR %% 2 == 0,
         "#9b59b6",
         "#d2b4de")
)


# GENE LABELS (MANHATTAN)

chr_centers <- combined_data %>%
  group_by(CHR) %>%
  summarise(chr_center = mean(plot_index)) %>%
  ungroup()

gene_labels_manhattan <- data.frame(
  gene = c("PER3", "HLA", "FADS1/2", "IRAK4", "SLC24A5"),
  CHR  = c(1, 6, 11, 12, 15),
  y    = c(-3.6, 4.5, -3.6, 4.5, -3.6)
) %>%
  left_join(chr_centers, by = "CHR") %>%
  rename(x = chr_center)


# MANHATTAN PLOT

manhattan_plot <- ggplot(combined_data,
                         aes(x = plot_index,
                             y = weighted_z,
                             color = color)) +
  geom_point(size = 0.7) +
  geom_hline(yintercept = c(3, -3),
             linetype = "dashed",
             color = "black") +

  annotate("text", x = max(combined_data$plot_index), y = 3.5,
           label = "z = 3",  hjust = 1.1, size = 3) +
  annotate("text", x = max(combined_data$plot_index), y = -3.5,
           label = "z = -3", hjust = 1.1, size = 3) +

  geom_text(data = gene_labels_manhattan %>% filter(gene != "HLA"),
            aes(x = x, y = y, label = gene),
            inherit.aes = FALSE,
            fontface = "italic",
            size = 2.5) +
  geom_text(data = gene_labels_manhattan %>% filter(gene == "HLA"),
            aes(x = x, y = y, label = gene),
            inherit.aes = FALSE,
            fontface = "plain",
            size = 2.5) +

  scale_color_identity() +
  labs(x = "Chromosome", y = "Average Z-score") +
  theme_minimal(base_size = 9) +
  theme(
    axis.text.x     = element_blank(),
    axis.ticks.x    = element_blank(),
    legend.position = "none",
    plot.margin     = margin(10, 10, 2, 10)
  )


# QQ PLOT

auto_qq <- combined_data %>%
  filter(CHR != 23, !is.na(p_value), p_value > 0)

x_qq <- combined_data %>%
  filter(CHR == 23, !is.na(p_value), p_value > 0)

qq_list <- list()

if (nrow(auto_qq) > 0) {
  qq_list[[1]] <- data.frame(
    expected = -log10(ppoints(nrow(auto_qq))),
    observed = -log10(sort(auto_qq$p_value)),
    group    = "Autosomes"
  )
}

if (nrow(x_qq) > 0) {
  qq_list[[2]] <- data.frame(
    expected = -log10(ppoints(nrow(x_qq))),
    observed = -log10(sort(x_qq$p_value)),
    group    = "ChrX"
  )
}

qq_combined <- bind_rows(qq_list)

qq_plot <- ggplot(qq_combined,
                  aes(x = expected,
                      y = observed,
                      color = group)) +
  geom_point(size = 0.6, alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0,
              color = "gray40",
              linetype = "dashed") +
  scale_color_manual(values = c(
    "Autosomes" = "#9b59b6",
    "ChrX"      = "#5dade2"
  )) +
  labs(x = "Expected",
       y = "Observed",
       color = NULL) +
  theme_minimal(base_size = 8) +
  theme(
    legend.position   = c(0.7, 0.2),
    legend.background = element_blank(),
    legend.key.size   = unit(1.2, "lines"),
    legend.text       = element_text(size = 6),
    plot.margin = margin(10, 10, 10, 10)  # was margin(10, 10, 2, 10)
  ) +
  guides(color = guide_legend(
    override.aes = list(size = 3)
  ))


# TOP PANEL (A + B)

top_row <- plot_grid(
  manhattan_plot,
  qq_plot,
  labels     = c("A", "B"),
  label_size = 14,
  rel_widths = c(2.5, 1.5)
)


# REPLICATION FUNCTION

make_chr_plot <- function(chrom, midpoint, show_legend = FALSE) {

  window_size <- 1e7

  combined <- bind_rows(
    z_score_data %>%
      filter(CHR == chrom & abs(POS - midpoint) < window_size) %>%
      mutate(dataset = "Discovery"),

    capture_z_score_data %>%
      filter(CHR == chrom & abs(POS - midpoint) < window_size) %>%
      mutate(dataset = "Replication 1"),

    rep_z_score_data %>%
      filter(CHR == chrom & abs(POS - midpoint) < window_size) %>%
      mutate(dataset = "Replication 2")
  )

  combined$POS_MB <- combined$POS / 1e6

  x_min <- (midpoint - window_size) / 1e6
  x_max <- (midpoint + window_size) / 1e6

  combined$dataset <- factor(combined$dataset,
                             levels = c("Discovery", "Replication 1", "Replication 2"))

  p <- ggplot(combined, aes(x = POS_MB, y = weighted_z, color = dataset)) +
    geom_point(size = 0.6) +

    geom_hline(yintercept = c(3, -3),
               linetype   = "dashed",
               color      = "black",
               linewidth  = 0.4) +

    geom_hline(yintercept = c(2.84, -2.84),
               linetype   = "solid",
               color      = "black",
               linewidth  = 0.4) +

    scale_color_manual(
      name   = "Dataset",
      values = c(
        "Discovery"     = "purple4",
        "Replication 1" = "darkgreen",
        "Replication 2" = "lightgreen"
      )
    ) +

    labs(x = "Position (Mb)", y = "Z-score") +
    coord_cartesian(xlim = c(x_min, x_max)) +
    theme_minimal(base_size = 7) +
    theme(
      legend.position = if (show_legend) "bottom" else "none",
      axis.text.x     = element_text(angle = 45, hjust = 1),
      plot.margin     = margin(5, 5, 5, 5)
    )

  # ---------- REGION BOX ----------
  gene_annotations <- data.frame(
    gene  = c("PER3", "HLA", "FADS1", "FADS2", "SLC24A5"),
    CHR   = c(1, 6, 11, 11, 15),
    start = c(7840000, 29600000, 61570000, 61620000, 48420000),
    end   = c(7902000, 33400000, 61590000, 61640000, 48440000)
  )

  pos_height <- c(2.8, 3.2)
  neg_height <- c(-3.2, -2.8)

  if (chrom == 6) {
    ann  <- gene_annotations %>% filter(CHR == 6)
    xmin <- ann$start / 1e6
    xmax <- ann$end   / 1e6
    ymin <- neg_height[1]
    ymax <- neg_height[2]
  } else {
    box_half_width <- 0.2
    xmin <- (midpoint / 1e6) - box_half_width
    xmax <- (midpoint / 1e6) + box_half_width

    if (chrom %in% c(1, 11, 15)) {
      ymin <- pos_height[1]; ymax <- pos_height[2]
    } else {
      ymin <- neg_height[1]; ymax <- neg_height[2]
    }
  }

  p <- p +
    annotate("rect",
             xmin = xmin, xmax = xmax,
             ymin = ymin, ymax = ymax,
             fill = "black")

  return(p)
}


# REPLICATION PANELS (C-G) — no legend

p1 <- make_chr_plot(1,  7063387)
p2 <- make_chr_plot(6,  30927867)
p3 <- make_chr_plot(11, 61906864)
p4 <- make_chr_plot(12, 44252015)
p5 <- make_chr_plot(15, 48433494)   # legend removed

rep_row <- plot_grid(
  p1, p2, p3, p4, p5,
  ncol       = 5,
  labels     = c("C", "D", "E", "F", "G"),
  label_size = 8,
  align      = "h"
)

# Gene labels above C-G
rep_gene_labels <- c("PER3", "HLA", "FADS1/2", "IRAK4", "SLC24A5")
x_positions     <- seq(0.1, 0.9, length.out = 5)

# x positions: just to the right of each panel letter
x_label_positions <- seq(0.05, 0.83, length.out = 5) 

rep_row_with_genes <- ggdraw(rep_row)
for (i in 1:5) {
  rep_row_with_genes <- rep_row_with_genes +
    draw_label(
      rep_gene_labels[i],
      x        = x_label_positions[i],
      y        = 0.97,
      hjust    = 0,          # left-align
      fontface = ifelse(rep_gene_labels[i] == "HLA", "plain", "italic"),
      size     = 7
    )
}



# ENRICHMENT DATA

methods <- c("ancestry_hmm", "mosaic", "rfmix", "simplai",
             "replication2", "replication1", "ancestry_paths")

full_paths <- c(
  "../final_figures_data/fig5_full_ancestry_hmm_average_ancestry_output.txt",
  "../final_figures_data/fig5_full_mosaic_average_ancestry_output.txt",
  "../final_figures_data/fig5_full_rfmix_average_ancestry_output.txt",
  "../final_figures_data/fig5_full_simplai_average_ancestry_output.txt",
  "../final_figures_data/fig5_full_replication2_average_ancestry_output.txt",
  "../final_figures_data/fig5_full_replication1_output_ancestry_hmm.txt",
  "../final_figures_data/fig5_full_ancestry_paths_average_ancestry_output.txt"
)

subset_paths <- c(
  "../final_figures_data/fig5_ancestry_hmm_average_ancestry_output.txt",
  "../final_figures_data/fig5_mosaic_average_ancestry_output.txt",
  "../final_figures_data/fig5_rfmix_average_ancestry_output.txt",
  "../final_figures_data/fig5_simplai_average_ancestry_output.txt",
  "../final_figures_data/fig5_replication2_average_ancestry_output.txt",
  "../final_figures_data/fig5_replication1_average_ancestry_output.txt",
  "../final_figures_data/fig5_ancestry_paths_average_ancestry_output.txt"
)

# Two-letter codes matching methods order
two_letter_codes <- c("AH", "Mo", "RF", "SI", "R2", "R1", "AP")

regions <- data.frame(
  gene  = c("HLA",    "IRAK4",  "FADS",   "SLC24A5", "PER3"),
  chr   = c(6,        12,       11,       15,         1),
  start = c(30000000, 42252015, 59000000, 46105416,   5063387),
  end   = c(33000000, 46252015, 65000000, 50105416,   9063387)
)

enrich_results <- data.frame()

for (r in 1:nrow(regions)) {
  region <- regions[r, ]

  for (i in seq_along(methods)) {
    method <- methods[i]

    subset_data     <- read.table(subset_paths[i], header = TRUE)
    overall_subset  <- mean(subset_data$HGall, na.rm = TRUE)
    region_subset   <- mean(subset_data$HGall[
      subset_data$CHR == region$chr &
      subset_data$POS > region$start &
      subset_data$POS < region$end
    ], na.rm = TRUE)
    subset_enrichment <- region_subset - overall_subset

    full_data       <- read.table(full_paths[i], header = TRUE)
    overall_full    <- mean(full_data$HGall, na.rm = TRUE)
    region_full     <- mean(full_data$HGall[
      full_data$CHR == region$chr &
      full_data$POS > region$start &
      full_data$POS < region$end
    ], na.rm = TRUE)
    full_enrichment <- region_full - overall_full

    enrich_results <- rbind(enrich_results, data.frame(
      gene       = region$gene,
      method     = method,
      short_code = two_letter_codes[i],
      type       = c("Full", "Subset"),
      enrichment = c(full_enrichment, subset_enrichment)
    ))
  }
}


# ENRICHMENT FACTOR ORDERING

enrich_results$gene <- factor(enrich_results$gene,
                               levels = c("PER3", "HLA", "FADS", "IRAK4", "SLC24A5"))

# Display order: 5 discovery methods then R1 then R2
display_order <- c("AH", "Mo", "RF", "SI", "AP", "R1", "R2")
enrich_results$short_code <- factor(enrich_results$short_code,
                                     levels = display_order)

# Dataset group drives colour
enrich_results$ds_group <- case_when(
  enrich_results$short_code %in% c("AH", "Mo", "RF", "SI", "AP") ~ "Discovery",
  enrich_results$short_code == "R1"                               ~ "Replication 1",
  enrich_results$short_code == "R2"                               ~ "Replication 2"
)
enrich_results$ds_group <- factor(enrich_results$ds_group,
                                   levels = c("Discovery", "Replication 1", "Replication 2"))

ds_colours <- c(
  "Discovery"     = "purple4",
  "Replication 1" = "darkgreen",
  "Replication 2" = "#90ee90"
)


# ENRICHMENT PLOTS (H-L)
# Full  = solid fill + matching outline, dodged left
# Subset = open bar (alpha=0 fill) + matching outline, dodged right

gene_order           <- c("PER3", "HLA", "FADS", "IRAK4", "SLC24A5")
panel_letters_enrich <- c("H", "I", "J", "K", "L")

enrich_plots <- list()

for (g in seq_along(gene_order)) {
  gname <- gene_order[g]
  dat_g <- enrich_results %>% filter(gene == gname)

  dat_full   <- dat_g %>% filter(type == "Full")
  dat_subset <- dat_g %>% filter(type == "Subset")

  # Assign explicit dodge group ids so both layers dodge together
  dat_full$dodge_id   <- paste0(as.character(dat_full$short_code),   "_F")
  dat_subset$dodge_id <- paste0(as.character(dat_subset$short_code), "_S")

dat_g$bar_fill <- ifelse(dat_g$type == "Full",
                            ds_colours[as.character(dat_g$ds_group)],
                            "white")

  ep <- ggplot(dat_g,
               aes(x = short_code, y = enrichment,
                   group = type)) +
    geom_bar(stat     = "identity",
             position = position_dodge(width = 0.6),
             width    = 0.5,
             linewidth = 0.4,
             aes(fill = bar_fill, colour = ds_group)) +
    scale_fill_identity() +
    scale_colour_manual(name = "Dataset", values = ds_colours) +
    geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4) +
    geom_vline(xintercept = 5.5, linetype = "dotted",
               linewidth = 0.5, color = "grey40") +
    labs(x = NULL, y = "HG Ancestry Enrichment") +
    theme_classic(base_size = 7) +
    theme(
      legend.position = "none",
      plot.title      = element_blank(),
      axis.text.x     = element_text(size = 6),
      plot.margin     = margin(18, 5, 5, 5)
    )

  enrich_plots[[g]] <- ep
}


# ENRICHMENT ROW + PANEL LETTERS (H-L via ggdraw)

enrich_row_raw <- plot_grid(
  plotlist = enrich_plots,
  ncol     = 5,
  align    = "h",
  axis     = "tb"
)

x_enrich <- seq(0.03, 0.83, length.out = 5) 

enrich_gene_labels <- c("PER3", "HLA", "FADS1/2", "IRAK4", "SLC24A5")

enrich_row <- ggdraw(enrich_row_raw)
for (g in 1:5) {
  enrich_row <- enrich_row +
    draw_label(
      panel_letters_enrich[g],
      x        = x_enrich[g],
      y        = 0.97,
      fontface = "bold",
      size     = 9
    ) +
    draw_label(
      enrich_gene_labels[g],
      x        = x_enrich[g] + 0.018,   # just right of the letter
      y        = 0.97,
      hjust    = 0,
      fontface = ifelse(enrich_gene_labels[g] == "HLA", "plain", "italic"),
      size     = 7
    )
}


# SHARED LEGEND FOR H-L
# Colour = dataset group; Fill style = Full (solid) vs Subset (open)

legend_helper <- data.frame(
  short_code = factor(rep(display_order, each = 2), levels = display_order),
  type       = rep(c("Full", "Subset"), length(display_order)),
  enrichment = 0.1
)
legend_helper$ds_group <- case_when(
  legend_helper$short_code %in% c("AH","Mo","RF","SI","AP") ~ "Discovery",
  legend_helper$short_code == "R1"                          ~ "Replication 1",
  legend_helper$short_code == "R2"                          ~ "Replication 2"
)
legend_helper$ds_group <- factor(legend_helper$ds_group,
                                  levels = c("Discovery","Replication 1","Replication 2"))
legend_helper$dodge_id <- paste0(legend_helper$short_code, "_",
                                  substr(legend_helper$type, 1, 1))

legend_enrich <- cowplot::get_legend(
  ggplot(legend_helper,
         aes(x = short_code, y = enrichment,
             fill = ds_group, colour = ds_group,
             alpha = type,
             group = dodge_id)) +
    geom_bar(stat = "identity",
             position = position_dodge2(width = 0.8, preserve = "single"),
             width = 0.35, linewidth = 0.5) +
    scale_fill_manual(
      name   = "Dataset",
      values = ds_colours
    ) +
    scale_colour_manual(
      name   = "Dataset",
      values = ds_colours
    ) +
    scale_alpha_manual(
      name   = "Sample set",
      values = c("Full" = 1, "Subset" = 0),
      guide  = guide_legend(
        override.aes = list(
          fill   = c("grey30", "white"),
          colour = c("grey30", "grey30"),
          alpha  = c(1, 1)
        )
      )
    ) +
    theme_void() +
    theme(
      legend.position = "bottom",
      legend.box      = "horizontal",
      legend.text     = element_text(size = 6),
      legend.title    = element_text(size = 6, face = "bold"),
      legend.key.size = unit(0.4, "lines")
    ) +
    guides(
      fill   = guide_legend(override.aes = list(alpha = 1)),
      colour = guide_legend(override.aes = list(alpha = 1))
    )
)


# ASSEMBLE FULL FIGURE

final_plot <- plot_grid(
  top_row,            # A + B
  rep_row_with_genes, # C–G  (no legend strip below)
  enrich_row,         # H–L
  legend_enrich,      # combined legend for H–L only
  ncol        = 1,
  rel_heights = c(1, 0.97, 0.62, 0.10)
)


# SAVE

ggsave("Figure5.pdf",
       final_plot,
       width  = 7,
       height = 5.5,
       dpi    = 300)
```

#Table 2 is put together from Table S1, see code for that Table for replication. 


